Large Language Model (LLM) agents face a double security vulnerability when using external tools with Model Context Protocols (MCPs): protocol-level attacks can bypass capability and source verification, while inference-level tampering can interfere with the agent's decision-making process by manipulating tool responses without triggering action-level monitoring.
Current defenses only address single-layer security: AttestMCP enhances the protocol but cannot detect the thought process involved in semantic tampering.
In contrast, guardrail models can only retrospectively evaluate actions and cannot verify the inference chains that generate them.
We propose \textsc{ReasoningGuard}, a dual-layer defense system integrating a Protocol-Attested Tool Gateway (PTG) and a Reasoning Trace Verifier (RTV).
On one hand, the PTG extends AttestMCP with signed intent digests, enabling the gateway to verify the authentication capabilities of each tool call before forwarding it.
On the other hand, the RTV captures the agent's externalized decision evidence into a structured trace and applies a constrained judge model to detect four classes of evidence inconsistency: Origin Awareness Violation (OAV), Intent-Action Divergence (IAD), Cross-server Action Injection (CAI), and Tool-Data Instruction (TDI)---a StruQ-inspired dimension that inspects invocation parameters for data exfiltration patterns and tool-response instruction following.
The theoretical basis of the dual-layer architecture is the observability-based defense limitation property: a defense mechanism at one architectural layer has significantly reduced detection power against attacks localized at another, because the primary attack signature lies outside the defender's observation space.
We evaluated \textsc{ReasoningGuard} on eight SOTA LLM agent backbones using our newly constructed MCPTox+ benchmark comprising 342 scenarios across six attack categories (Tool Description Poisoning, Parameter Injection, Response Manipulation, Capability Escalation, Context-Dependent, and Cross-Session Memory Poisoning).
Across 9{,}192 evaluation episodes, \textsc{ReasoningGuard} consistently reduces the Attack Success Rate from 10.8\%--74.2\% (No Defense) to 0.8\%--29.2\%, achieving a relative reduction of 48.5\%--98.9\% across all eight models.
On GPT-4o, the most comprehensively evaluated model with the improved four-dimensional RTV, \textsc{ReasoningGuard} reduces ASR from 74.2\% to 0.8\% (98.9\% relative reduction), achieves 0.0\% L2-ASR, and maintains 60.0\% TCR with 926\,ms defense overhead per tool call (LLM judge).
Ablation studies across three models confirm that intent attestation is the most critical component (removing it increases ASR by 6.7--18.3\,pp), PTG is essential for L4 defense, and RTV provides clear incremental value on L2 attacks (reducing L2 ASR to 0.0\% beyond PTG alone).
Multi-seed evaluation (3 seeds) with paired $t$-tests confirms statistical significance ($p<0.01$) on 3/4 models.
Our code and data are available at \url{https://github.com/939386360/ReasoningGuard}.
