\section{Conclusion}

We presented \textsc{ReasoningGuard}, a dual-layer defense framework for MCP-based LLM agents that combines Protocol-Attested Tool Gateway (PTG) with Reasoning Trace Verifier (RTV). Grounded in the observability-based defense limitation property, our approach addresses both protocol-level (L4) and cognitive-level (L2) attacks that no single-layer defense can comprehensively cover. PTG extends AttestMCP with semantic intent attestation, providing an intent-aware capability boundary that ablation identifies as the most critical single defense component. RTV provides evidence consistency checking with four anomaly classes (OAV, IAD, CAI, TDI), achieving 100\% TPR and 0\% FPR on direct evaluation. The TDI dimension's invocation-parameter inspection---detecting external email addresses, method-level intent divergence, and tool-response instruction following---effectively closes the L2 gap, reducing L2-ASR to 0.0\% on GPT-4o. We position RTV as an evidence consistency checker rather than a reasoning correctness verifier, acknowledging that LLM-generated traces are post-hoc justifications that may not faithfully reflect internal computation. Across eight SOTA models and 9{,}192 evaluation episodes, \textsc{ReasoningGuard} consistently achieves the lowest ASR, reducing it by 48.5\%--98.9\% relative to No Defense. On GPT-4o with the improved RTV, ASR drops to 0.8\% (98.9\% relative reduction) with 0.0\% L2-ASR, comprehensively outperforming all prompt-based and graph-based baselines on security. We identify adaptive adversary evasion strategies and discuss their feasibility. Our MCPTox+ benchmark of 342 scenarios across six attack categories provides the first comprehensive evaluation of cross-session (T3) memory poisoning in MCP security. Future work should explore neural judge models for adaptive adversary resistance, formal authorization integration, multi-seed validation of the improved RTV across additional models, and extension to multi-agent architectures.
